---
title: "Project1"
author: "Ailin Wu"
date: "2023-09-28"
output: pdf_document
---
The data in the file “UN.txt” contains PPgdp, the 2001 gross national product per
person in US dollars, and Fertility, the birth rate per 1000 femals in the population in the year 2000. The
data are for 184 localities, mostly UN member countries, but also other areas such as Hong Kong that are
not independent countries. In this problem, we study the relationship between Fertility and PPgdp.

#Data visualization and pre-processing
1. Draw the scatterplot of Fertility on the vertical axis versus PPgdp on the horizontal axis and summarize
the information in this graph. Does a simple linear regression model seem to be a plausible for a
summary of this graph?
```{r}
UNdata <- read.table ("/Users/ailin3/Downloads/UN.txt", header = TRUE, 
                        stringsAsFactors = FALSE)
head(UNdata)
summary(UNdata)
library(ggplot2)
ggplot(UNdata, aes(x = PPgdp, y = Fertility)) +
  geom_point(color = "blue") +
  geom_smooth(color = "red") +
  labs(x = "PPgdp", y = "Fertility", 
       title = "Fertility vs. Gross National Product Per Person USD")

```
#I think a simple linear regression model seem to be not a plausible for a summary of this graph because the mean is not linear and the variance is not constant. There is no linear relationship between the two variables.

2. In order to get a better fit, we seek to transform the variables. What transformations you would take
so that a simple linear regression model is proper? State why you choose these transformations. Draw
the scatter plot of the transformed variables. Comment on the plot.
```{r}
ggplot(UNdata, aes(x = log(PPgdp), y = log(Fertility))) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", color = "red") +
  labs(x = "PPgdp", y = "Fertility", 
       title = "Ntural Log of Fertility vs. Gross National Product Per Person USD")

```
#I took natural logarithms of both of the variables to make the simple linear regression model proper. I chose to use natural logarithms for the transformation because the x and y axis would be determined if there are any plausibility indication of a straight line mean function. The scatter plot after the transformation of the natural logarithmic appears to be more effective of a linear regression. This plot is more linear and the variance is more plausible than the previous plot. In conclusion, there is a linear relationship between fertility and ppgdp.


#Model fitting and diagnostics
3. Fit the simple linear model on the transformed data through three ways. Report the least square
estimates for the coefficients and R2. Add the fitted line to the scatter plot on the transformed data
and comment on the fit.
(a) Plain coding (not using the ‘lm’ function or matrix manipulation)
(b) Using the ‘lm’ function
(c) Through matrix manipulation
```{r}
#(a) Plain coding (not using the ‘lm’ function or matrix manipulation)
x <- log(UNdata$PPgdp)
y <- log(UNdata$Fertility)

ssx <- sum((x-mean(x))^2)

beta1 <- (sum((x-mean(x))*(y-mean(y)))/ssx)
beta1
beta0 <- mean(y)-(beta1*mean(x))
beta0

ssr <- ((beta1)^2) * ssx

sst <- sum((y - mean(y))^2)

R2 <- ssr/sst #what, it should be 0.2658
R2

UNdata$logFertility = log(UNdata$Fertility)
UNdata$logPPgdp = log(UNdata$PPgdp)


#(b) Using the ‘lm’ function
model <- lm(logFertility ~ logPPgdp,  UNdata)
summary(model)
beta0 <- coef(model) [1]
beta0
beta1 <- coef(model) [2]
beta1


#how to do r 2
model_sum = summary(model)
model_sum$r.squared

UNdata$resid = residuals(model)
#(c) Through matrix manipulation
library(MASS)

x0 = rep(1, nrow(UNdata))
x1 = x

X = cbind(x0, x1)

betahat = ginv(t(X)%*% X) %*% t(X) %*% y
betahat

#r2
#r2 = 1- sse/sst

#scatterplot
ggplot(UNdata, aes(x = log(PPgdp), y = log(Fertility))) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", color = "red") +
  labs(x = "PPgdp", y = "Fertility", 
       title = "Ntural Log of Fertility vs. Gross National Product Per Person USD")
```
# For the scatter plot, it has a negative correlation between log of fertility and log of PPgdp. The strength of the correlation is not really strong or weak. The trend appears to be more linear. 


4. Draw the diagnostic plots and comment
```{r}
#plot(model)

gg1 = ggplot(UNdata, aes(x = logPPgdp, y = logFertility)) + 
  geom_point()



gg2 = ggplot(UNdata, aes(sample = resid))
gg2 = gg2 + stat_qq() + stat_qq_line()


gg3 = ggplot(UNdata, aes(x = resid)) + 
  geom_histogram(bins = 20, color = "black", fill = "white") + 
  xlim(c(-1,1))


gg4 = ggplot(UNdata) + 
  geom_point(aes(x = logPPgdp, y = resid)) + 
  geom_hline(aes(yintercept = 0), color = "red")

library(ggpubr)
ggarrange(gg1, gg2, gg3, gg4, nrow = 2, ncol = 2)
```
# For the histogram plot is roughly normal, but something i would be worried about is the heavy left tail. 
# For the Normal QQ plot, the observation quantile is  under the theoretical line, so the distribution of residual is heavy tailed the inference is not good/under powered. 
# For the scatter plot shows linear relationship between X an Y.
# For the residual plot, there is no pattern, so its okay. This is a homeostatic plot of residuals, which implies our assumption of constant variance is right. 

#Inference
5. Test whether there is a linear relationship between the transformed variables.
```{r}
#f-test
n <- length(x)
yhat <- model$fitted.values
sse <- sum((y - (yhat))^2)
mse <- sse/(n-2)

ftest <- ssr/mse

Pvalue <- 1 - pf(abs(ftest), 1, n-2) #right tail
alpha <- 0.05

if (Pvalue < alpha) {
  cat("Reject the null hypothesis. There is a linear association between 
      between the transformed variables. ")
} else {
  cat ("Fail to reject the null hypothesis. There is no suffficient evidence 
       of a linear association.")
}

```


6. Provide a 99% confidence interval on the expected Fertility for a region with PPgdp 20,000 US dollars
in 2001.
```{r}
alpha <- 0.01

#E(Y|X=20000)
xh <- log(20000)
yhatx <- beta0 + beta1 * x

t <- qt(0.995, n-2)
t

yhat = function(xh){
  beta0+beta1*xh
}

SEyhat = function(xh) {
  sqrt(mse * (1/n + (xh - mean(x))^2/ssx))
}

lowerbound <- yhat(xh) - t * SEyhat(xh)
uppderbound<- yhat(xh) + t * SEyhat(xh)
cat(lowerbound, uppderbound) # 99% CI for log(E(Y|X=20000))
exp(c(lowerbound, uppderbound)) #undoing the log


#built in function 
new_df = data.frame(logPPgdp = log(20000))
ci = predict(model, new_df, interval = "confidence", level = 0.99)
exp(ci)
```



7. Provide a 95% confidence band for the relation between the expected Fertility and PPgdp. Add the
bands to the scatter plot of the original data.
```{r}

#WH cv
xs = seq(from = range(UNdata$logPPgdp)[1], to = range(UNdata$logPPgdp)[2], length.out =100)

WH = function(xs) {
  mat = matrix(0, nrow = length(xs), ncol = 3)
  mat[,1 ] = xs
  for(i in 1:length(xs)){
    mat[i,2:3] = yhat(xs[i]) +c(-1,1)* sqrt(2*qf(0.95, 2, n-2))*SEyhat(xs[i])
  }
  return(mat)
}

df = WH(xs)
colnames(df) = c("x", "lower", "upper")
df = as.data.frame(df)

#log plot
ggplot(UNdata, aes(x = log(PPgdp), y = log(Fertility))) +
  geom_point(color = "blue") +
  stat_smooth(method = "lm", color = "red", level = 0.95) +
  labs(x = "PPgdp", y = "Fertility", 
       title = "Ntural Log of Fertility vs. Gross National Product Per Person USD") +
  geom_line(data = df, mapping = aes(x = x, y = lower), color = "red", linetype = "dashed")+
  geom_line(data = df, mapping = aes(x = x, y = upper), color = "red", linetype = "dashed")

#original plot
ggplot(UNdata, aes(x = PPgdp, y = Fertility)) +
  geom_point(color = "blue") +
  #stat_smooth(method = "lm", color = "red", level = 0.95) +
  labs(x = "PPgdp", y = "Fertility", 
       title = "Confidence Band of Mean Fertility vs. Gross National Product Per Person USD") +
  geom_line(data = df, mapping = aes(x = exp(x), y = exp(lower)), color = "red", linetype = "dashed")+
  geom_line(data = df, mapping = aes(x = exp(x), y = exp(upper)), color = "red", linetype = "dashed")

```



8. Assuming that the same relationship between Fertility and PPgdp holds, give a 99% prediction interval
on Fertility for a region with PPgdp 25,000 US dollars in 2018.
```{r}
alpha <- 0.01

#E(Y|X=20000)
xh <- log(25000)
yhatx <- beta0 + beta1 * x

t <- qt(0.995, n-2)
t

yhat = function(xh){
  beta0+beta1*xh
}

SEyhat = function(xh) {
  sqrt(mse * (1 + 1/n + (xh - mean(x))^2/ssx))
}

lowerbound <- yhat(xh) - t * SEyhat(xh)
uppderbound<- yhat(xh) + t * SEyhat(xh)
cat(lowerbound, uppderbound) # 99% CI for log(E(Y|X=20000))
exp(c(lowerbound, uppderbound)) #undoing the log


#built in function 
new_df = data.frame(logPPgdp = log(25000))
pi = predict(model, new_df, interval = "prediction", level = 0.99)
exp(pi)
```



9. Based on the diagnostic plots in Part 4, do you have any concern on the above hypothesis testing and
inferences? If so, what are the concerns?
#For the above hypothesis testing and inferences, I would conclude in the QQ plot that the true coverage would be less because of the heavy tail, in the scatter plot there is a linear relationship between PPgdp and Fertility, in the residual plot there is no pattern so our assumption of linearity and the scatter width is constandt so residuals are homoskedastic, and the histogram shows it is roughly a normal distribution with a heavy tail.



